---
title: "Setup"
output: html_document
date: "2025-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Make sure to run "install.packages("renv")" in the console, and then run "renv::restore()" to retrieve the packages needed to run these markdown files - this ensures that we are all using the same packages

```{r, eval = FALSE}
library(tidycensus)
library(knitr)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r}
Sys.getenv("CENSUS_API_KEY")
```

```{r}
#Run this chunk to view the vars within tidycensus 
viewvars <- load_variables(2023, "acs5", cache = TRUE)
view(viewvars)
```


#Testing
```{r}
#load in the needed variables from tidy census 
texas_total_pop <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_001",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
ggplot(texas_total_pop, aes(fill = B02001_001E)) + 
  geom_sf() + 
  theme_void() +
  scale_fill_viridis_c(option = "turbo") 
```

```{r}
texas_food_stamps <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B22003_002",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
ggplot(texas_food_stamps, aes(fill = B22003_002E)) + 
  geom_sf() + 
  theme_void() +
  scale_fill_viridis_c(option = "turbo") 
```

```{r}
houston_pop <- get_acs(
  state = "Texas",
  county = "Harris County",
  geography = "tract",
  variables = "B02001_001",
  geometry = TRUE,
  output = "wide",
  year = 2022)

# c("Austin County", "Brazoria County", "Chambers County", "Fort Bend County", "Galveston County", "Harris County", "Liberty County", "Montgomery County", "Waller County")
```


```{r}
ggplot(houston_pop, aes(fill = B02001_001E), (size = 0.00001)) + 
  geom_sf() + 
  theme_void() +
  scale_fill_viridis_c(option = "turbo") 
```

```{r}
houston_food_stamps <- get_acs(
  state = "Texas",
  county = "Harris County",
  geography = "tract",
  variables = "B22003_002",
  geometry = TRUE,
  output = "wide",
  year = 2022)

# c("Austin County", "Brazoria County", "Chambers County", "Fort Bend County", "Galveston County", "Harris County", "Liberty County", "Montgomery County", "Waller County")
```

```{r}
ggplot(houston_food_stamps, aes(fill = B22003_002E), (size = 0.00001)) + 
  geom_sf() + 
  theme_void() +
  scale_fill_viridis_c(option = "turbo") 
```

#Create Dataframe

##Load in the data: 

```{r}
food_stamps <- get_acs(
 state = "Texas",
  geography = "county",
  variables = "B22003_002",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
white <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_002",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
black<- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_003",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
asian <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_005",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
american_indian_alaska_native <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_004",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
pacific_islander <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_006",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
other_race <- get_acs(
  state = "Texas",
  geography = "county",
  variables = "B02001_007",
  geometry = TRUE,
  output = "wide",
  year = 2022)
```

```{r}
#Rename columns for simplicity 
names(food_stamps)[3] <- "food_stamp"
names(white)[3] <- "white"
names(black)[3] <- "black"
names(asian)[3] <- "asian"
names(american_indian_alaska_native)[3] <- "american_indian"
names(pacific_islander)[3] <- "pacific_islander"
names(other_race)[3] <- "other"
```

##Merge
```{r}
#Merge
df_race <- white |>
mutate(     
    black = black$black,
    asian = asian$asian,
    american_indian = american_indian_alaska_native$american_indian,
    pacific_islander = pacific_islander$pacific_islander,
    other = other_race$other, 
    food_stamp = food_stamps$food_stamp)

```

```{r}
#remove uneeded column 
df_race <- df_race[, -4]
```

```{r}
#create a total population column
df_race <- df_race |>
  mutate(
    total_pop = (black + asian + white + american_indian + pacific_islander + other)
  )
```

##Per capita populations

```{r}
df_race <- df_race |>
  mutate(
    white_p = white/total_pop,
    black_p = black/total_pop,
    asian_p = asian/total_pop,
    american_indian_p = american_indian/total_pop,
    pacific_islander_p = pacific_islander/total_pop,
    other_p = other/total_pop,
    food_stamp_p = food_stamp/total_pop
)
```

```{r}
#save
saveRDS(df_race, file = "df_race.rds")
```

#To Evoke: 
```{r}
readRDS("df_race.rds")
```


Tidycensus variables: 

Total population
B02001_001 (or B01003_001)
White population
B02001_002
Black population
B02001_003
Asian population
B02001_005
American Indian and Alaska Native population
B02001_004
Native Hawaiian or other pacific islander population
B02001_006
Some other race: 
B02001_007

Exclusions: Two or more races