---
title: "Setup"
output: html_document
date: "2025-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Make sure to run "install.packages("renv")" in the console, and then run "renv::restore()" to retrieve the packages needed to run these markdown files - this ensures that we are all using the same packages

```{r, eval = FALSE}
library(tidycensus)
library(knitr)
library(tidyverse)
library(ggplot2)
```

```{r}
Sys.getenv("CENSUS_API_KEY")
```

```{r}
#Run this chunk to view the vars within tidycensus 
viewvars <- load_variables(2023, "acs5", cache = TRUE)
view(viewvars)
```

```{r}
#load in the needed variables from tidy census 
#

```

## R Markdown

Tidycensus variables to remember:

Total population
B02001_001 (or B01003_001)
White population
B02001_002
Black population
B02001_003
Asian population
B02001_005
American Indian and Alaska Native population
B02001_004
Native Hawaiian or other pacific islander population
B02001_006
Latino population
B03002_012